Sara Benist MADA Project

### 2023-03-17 code/analysis_code/exploratory_analysis.qmd
Part 3 of project
beginning of statistical analysis, model fitting, update readme files, highburden subset

To start, the needed packages are loaded and the `processeddata.rda` is loaded from the `processed_data` folder.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup
The `exploratoyranalysis.r` script needs to be indicated as the file from which to pull code chunks from. 
```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('exploratoryanalysis.R')
```

@@

For each indicator, I will be creating a plot using `ggplot()` to study differences between subgroups. For most of these variables, I will be using a boxplot in order to compare the summary measures between subgroups. For viewing multiple dimensions and comparing subgroups within each dimension, I will be colors and grouping through `ggplot()`.

@@

The forth plot shows drug resistance by sex. Neither group appears to have a higher percentage of drug resistance.
```{r, exploratoryfigures3}
```
The fifth plot and sixth plot shows TB incidence and mortality rates by country, separated by sex. The countries are ordered from lowest to highest rate, and males seem to have a higher proportion of TB cases.

@@

The eighth plot shows TB prevalence by country, showing differences between rural and urban areas. There appears to be a small trend of urban areas having higher prevalence then rural. 
```{r, exploratoryfigures4}
```
The ninth plot shows the overall attitude towards keeping a family member's TB a secret by sex. Females tend to be correlated with a higher percentage compared to males.

@@

In the next proportion of the exploratory analysis, I would like to examine more closely the influence of outliers (and if the data needs to be separated by burden of TB) as well as start comparing indicators to TB mortality, incidence, and prevalence.

>>>>

To start, the needed packages are loaded, and the `processeddata.rda` is loaded from the `processed_data` folder.

# Setup
The `exploratoryanalysis.r` script needs to be indicated as the file from which to pull code chunks from. 
```{r, include=FALSE, cache=FALSE}
knitr::read_chunk('exploratoryanalysis.R')
```

@@

For each indicator, I will be creating a plot using the `highburden` data set and `ggplot()` to study differences between subgroups. For most of these variables, I will be using a boxplot in order to compare the summary measures between subgroups. For viewing multiple dimensions and comparing subgroups within each dimension, I will be colors and grouping through `ggplot()`.

@@

The forth plot shows drug resistance by sex. Neither group appears to have a higher percentage of drug resistance.
```{r, exploratoryfigures3}
```
The 5a plot and 5b plots show TB incidence by country, separated by sex for the high burden countries and full data set. The countries are ordered from lowest to highest rate, and males seem to have a higher proportion of TB cases.The `highburden` data set appears to include most of the countries that have at least one new case per 100,000.

The sixth plot shows TB mortality by country, separated by sex. The countries are ordered from lowest to highest rate, and males tend to have a mortality rate.

@@

The 8a and 8b plots show TB prevalence by country, showing differences between rural and urban areas for the `highburden` and full data set. There appears to be a small trend of urban areas having higher prevalence then rural. 
```{r, exploratoryfigures4}
```
The ninth plot shows the overall attitude towards keeping a family member's TB a secret by sex. Females tend to be correlated with a higher percentage compared to males.

@@

In the next proportion of the exploratory analysis, I would like to examine more closely the influence of outliers as well as start comparing indicators to TB mortality, incidence, and prevalence.

### 2023-03-17 code/processing_code/processingfile.qmd
Part 3 of project
beginning of statistical analysis, model fitting, update readme files, highburden subset

```{r, exploredata1}
```
The data has the countries listed under setting, year of the data, the indicator that is being studied, and the dimension and subgroup for each observation. The estimate column holds the actual percentage or number of interest. Other variables on the data set contain formatting information.

@@

I would also like a list of the indicator names and a count of how many countries are in the data set. 
```{r, exploredata2}
```
There are 14 indicators and 194 countries.

@@

```{r, cleandata1}
```
The `wide_data` had 7473 rows and 20 columns.

## Subset by indicator
Next, I would like to create new objects for each indicator so I can make graphs/figures later on. I will use the `filter()` and `select()` functions to subset each indicator and reassign to new objects.

Here, the BCG coverage, drug resistance, and case detection rate indicators are created. 
```{r, cleandata2}
```

Here, the percentage reporting TB is spread through cough, overall and by sex, indicators are created.

@@

Here, the TB incidence and TB mortality indicators are created.

@@

Here, the percentage wanting to keep TB infection a secret, overall and by sex, indicators are created.

@@

Here, the TB prevalence and prevalance to notification ratio indicators are created.

@@

Finally, the percentage of afftected people suffering catastrophic costs due to TB indicator is created. 
```{r, cleandata7}
```

All of these indicator subsets can now be individually graphed and the trends can be compared. Most of the data seems to be from 2004-2020, and the estimates are within the bounds of the indicator (such as all the percentages are between 0 and 100).

@@

Looking over the data more closely, the missing data (marked as `NA`) is due to a country and year having data for only one subgroup in the dimension. We can drop these observations without losing important data analysis opportunities.

```{r, cleandata8}
```

@@

The prevalence indicator and prevalence to notification indicator both have no population data associated with the variable. I will drop the population column for both. Catastrophic cost also is missing a lot of data for the population column, but since there is some data available for a few observations, I will leave the indicator as is.

```{r, cleandata9}
```

>>>>

```{r, exploredata1}
```

The data has the countries listed under setting, year of the data, the indicator that is being studied, and the dimension and subgroup for each observation. The estimate column holds the actual percentage or number of interest. Other variables on the data set contain formatting information.

@@

I would also like a list of the indicator names and a count of how many countries are in the data set.

```{r, exploredata2}
```

There are 14 indicators and 194 countries.

@@

I will also be looking specifically at high TB burden countries. The high burden countries list is updated by the WHO and can be found [here](https://www.who.int/news/item/17-06-2021-who-releases-new-global-lists-of-high-burden-countries-for-tb-hiv-associated-tb-and-drug-resistant-tb). I will subset these countries in the `highburden` object.

```{r, cleandata1}
```

The `wide_data` had 7473 rows and 20 columns. The `highburden` object has 3761 rows and 20 columns.

## Subset by indicator

Next, I would like to check each indicator for potential data cleaning issues I need to address. I will use the `filter()` and `select()` functions to subset each indicator and `summary()` to check each indicator.

Here, the BCG coverage, drug resistance, and case detection rate indicators are checked.

```{r, cleandata2}
```

Here, the percentage reporting TB is spread through cough, overall and by sex, indicators are checked.

@@

Here, the TB incidence and TB mortality indicators are checked.

@@

Here, the percentage wanting to keep TB infection a secret, overall and by sex, indicators are checked.

@@

Here, the TB prevalence and prevalence to notification ratio indicators are checked.

@@

Finally, the percentage of affected people suffering catastrophic costs due to TB indicator is checked.

```{r, cleandata7}
```

Most of the data seems to be from 2004-2020, and the estimates are within the bounds of the indicator (such as all the percentages are between 0 and 100).

@@

Looking over the data more closely, the missing data (marked as `NA`) is due to a country and year having data for only one subgroup in the dimension. Using the `drop_na()` function would drop all of the data since every row is incomplete. For now, I won't change the data until I have explored more.

@@

### 2023-03-17 products/manuscript/Manuscript.qmd
Part 3 of project
beginning of statistical analysis, model fitting, update readme files, highburden subset

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

@@

The `rawdata` file contains information on the country, year of data collection, the indicator they are studying, the social determinants of health they are considering, and the subgroup of each dimension. The dataset is mostly complete with only a few main variables missing data points
```{r}
#another view of the data
skimr::skim(rawdata)
```

The indicators being studied are shown below.

@@

For the main part of the data cleaning, I used `pivot_wider()` to allow each indicator to have its own column, and assign each indicator into its own object using `filter()` and `select()` functions. These methods will allow for deeper exploration of each indicator in order to explore trends and patterns for each subgroup. An example of the subsetting is shown below. 
```{r}
wide_data <- rawdata %>% 
  select(c(setting, year, indicator_name,
           indicator_abbr, dimension, subgroup,
           estimate, population)) %>% 
  pivot_wider(names_from = "indicator_name", values_from = "estimate")
slice(wide_data)

# BCG coverage indicator
bcg <- wide_data %>% 
  filter(indicator_abbr == "bcg") %>% 
  select(c(1,2,3,4,5,6,7))
summary(bcg)
```

Due to the `pivot_wider()` function, there was several missing observations for countries and years that had data for only some subgroups (ex: data on males but no data on females). Most of these observations were dropped since no important information was lost by removing these "empty" observations. An example is shown below.
```{r}
# BCG missing population & bcg data 
# due to missing info for dimension levels
bcg <- bcg %>% 
  drop_na()
```

The processed and cleaned data was then saved into the `processed_data` folder under `processeddata.rda` to be used during the exploratory analysis phase.

## Statistical analysis

TBA: statistical analysis description

@@

```{r}
#incidence summary and save to file location
summary_inc <- skimr::skim(incidence)
print(summary_inc)
incsummarytable_file <- here("results", "summary tables", "incsummarytable.rds")
saveRDS(summary_inc, file = incsummarytable_file)

#mortality summary and save to file location
summary_mort <- skimr::skim(mortality)
print(summary_mort)
mortsummarytable_file <- here("results", "summary tables", "mortsummarytable.rds")
saveRDS(summary_mort, file = mortsummarytable_file)
```
The population for most variables were relatively low and skewed to the right. The TB incidence and mortality indicators also showed a skewed distribution. The highest incidence rate was 908 cases/100,000 and highest mortality rate was 140 deaths/100,000.

@@

```{r}
p13 <- tb_cough_f %>% ggplot(aes(x=dimension,
                                 y = `People who report TB is spread through coughing - Female (%)`,
                                 fill = subgroup)) +
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90))   
plot(p13) # plots knowledge about TB by dimension and subgroup for females
```

The main outcomes of interest for TB incidence, mortality, and prevalence were plotted using the `geom_point()` function as shown below.
```{r}
p6 <- mortality %>% ggplot(aes(x=fct_reorder(
  setting, `TB mortality (deaths per 100 000 population)`),
  y = `TB mortality (deaths per 100 000 population)`,
  color = subgroup)) +
    geom_point()+
  theme(axis.text.x = element_text(angle = 90))+
  scale_fill_brewer(palette = "Spectral")+
  xlab("setting")
plot(p6) #plots mortality by country from lowest mortality to highest, colored by subgroup
```

@@

## Basic statistical analysis
TBA: simple statistical models for associations in data

## Full analysis
TBA: final analysis with statistic/ML methods, meaningful figures/tables, load saved products here

>>>>

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(tidymodels)
```

@@

The `rawdata` file contains information on the country, year of data collection, the indicator they are studying, the social determinants of health they are considering, and the subgroup of each dimension. The dataset is mostly complete with only a few main variables missing data points. The indicators being studied are shown below.

@@

For the main part of the data cleaning, I used `pivot_wider()` to allow each indicator to have its own column, and subset the high burden countries (as indicated by the WHO) into a new object called `highburden`. I also checked each indicator using `filter()`, `select()`, and `summary()` functions. These methods will allow for deeper exploration of each indicator in order to explore trends and patterns for each subgroup. An example of the data cleaning is shown below. 
```{r}
wide_data <- rawdata %>% 
  select(c(setting, year, indicator_name,
           indicator_abbr, dimension, subgroup,
           estimate, population)) %>% 
  pivot_wider(names_from = "indicator_name", values_from = "estimate")
head(wide_data)

# BCG coverage indicator
wide_data %>% 
  filter(indicator_abbr == "bcg") %>% 
  select(c(1,2,3,4,5,6,7)) %>% 
  summary()
```

Due to the `pivot_wider()` function, there was several missing observations for countries and years that had data for only some subgroups (ex: data on males but no data on females). Since I will be maintaining all the indicators in their own data set, `drop_na()` would remove all data points. Instead, I will handle missing data points after exploring the data further.

The processed and cleaned data was then saved into the `processed_data` folder under `processeddata.rda` to be used during the exploratory analysis phase. I have loaded the processed data below.
```{r}
#path to data
data_location <- here::here("data","processed_data","processeddata.rda")
#load data. 
load(data_location)
```

## Statistical analysis
*(Part 3 of project: statistical analysis)*
For the statistical analysis of the data, I created linear models for the main outcomes (TB incidence, TB mortality, and TB prevalence) based on the subgroups of the indicator. An example is shown below. This method was repeated for the `wide_data` and `highburden` countries.

```{r}
# fit linear model using TB incidence as outcome, subgroup as predictor
lm_mod <- linear_reg() #set linear regression

#fit model with high burden countries
inclm_fit <- lm_mod %>% 
  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup,
      data = highburden) #fit linear model
incfittable <- tidy(inclm_fit) 
print(incfittable) #produce tidy table of fitted model
```

Other indicators were also analyzed and fit to a linear model as shown below. 
```{r}
# fit linear model with BCG coverage as outcome, subgroups as predictors

#fit model with high burden countries
BCGlm_fit <- lm_mod %>% 
  fit(`BCG immunization coverage among one-year-olds (%)` ~ subgroup,
      data = highburden) #fit linear model
BCGfittable <- tidy(BCGlm_fit) 
print(BCGfittable) #produce tidy table of fitted model
```

In the next step of my project, I would like to manipulate the data set so I can use the social determinant indicators as predictors for the main outcomes. This will most likely be limited to the high burden countries.

@@

```{r}
#incidence summary and save to file location
summary_inc <- skimr::skim(wide_data$`TB incidence (new infections per 100 000 population)`)
print(summary_inc)

#mortality summary and save to file location
summary_mort <- skimr::skim(wide_data$`TB mortality (deaths per 100 000 population)`)
print(summary_mort)
```
The population for most variables were relatively low and skewed to the right. The TB incidence and mortality indicators also showed a skewed distribution. The highest incidence rate was 908 cases/100,000 and highest mortality rate was 140 deaths/100,000.

@@

```{r}
p13 <- highburden %>%
  filter(!dimension %in% c("Age (2 groups) (0-15+)", "TB drug resistance", "Sex")) %>%
  ggplot(aes(x=dimension,
             y = `People who report TB is spread through coughing - Female (%)`,
             fill = subgroup)) +
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90))   
plot(p13) # plots knowledge about TB by dimension and subgroup for females
```

The main outcomes of interest for TB incidence, mortality, and prevalence were plotted using the `geom_point()` function as shown below.
```{r}
p6 <- highburden %>% 
  filter(dimension %in% "Sex") %>%
  filter(indicator_abbr == "mortality") %>% 
  select(c(1,2,3,4,5,6,13)) %>% 
  ggplot(aes(x=fct_reorder(
  setting, `TB mortality (deaths per 100 000 population)`),
  y = `TB mortality (deaths per 100 000 population)`,
  color = subgroup)) +
    geom_point()+
  theme(axis.text.x = element_text(angle = 90))+
  scale_fill_brewer(palette = "Spectral")+
  xlab("setting")
plot(p6) #plots mortality by country from lowest mortality to highest, colored by subgroup
```

@@

## Basic statistical analysis
*(Part 3 of project: linear model fitting and predictions, more to be added)*
TBA: simple statistical models for associations in data

The linear models summary for TB incidence are listed below.The high burden model has the linear equation of `TB incidence = 194 + 126(Male)` where male would equal 1. The full data model has the linear equation of `TB incidence = 82.6 + 48(Male)`. Based on the p-value for the subgroup variable, sex has a significant association with TB incidence in both models. Males having higher TB incidence is congruent with the exploratory analysis findings. The full model has smaller coefficients compared to `highburden` model. 
```{r}
readRDS(file = here("results", "model fit tables", "incfittable.rds")) #TB incidence model for high burden countries
readRDS(file = here("results", "model fit tables", "incfittableFD.rds")) #TB incidence model for full data
```
The fitted model summaries below are the TB mortality for high burden countries and full data. For high burden countries, the linear equation is `TB mortality = 23.6 + 14.9(Male)`; for the full data model, the linear equation is `TB mortality = 9.98 + 6.43(Male)`. Sex is a statistically significant predictor of TB mortality (p-values <0.05), but this does not consider any other factors. The next step in my analysis would be to rearrange the data set so I can predict the main outcomes using the sociodemographic indicators. 
```{r}
readRDS(file = here("results", "model fit tables", "mortfittable.rds")) #TB mortality model for high burden countries
readRDS(file = here("results", "model fit tables", "mortfittableFD.rds")) #TB mortality model for full data
```
The last outcome model I fitted was TB prevalence for high burden countries and the full data. For high burden countries, the linear equation is `TB Prevalence = 399 + 151(Urban)` with urban residences receiving a 1 and rural residences receiving a 0. The full data model has an equation of `TB Prevalence = 410 + 95.5(Urban)`. 
```{r}
readRDS(file = here("results", "model fit tables", "prevfittable.rds")) #TB prevalence model for high burden countries
readRDS(file = here("results", "model fit tables", "prevfittableFD.rds")) #TB prevalence model for full data
```
The models can be further explored by including `setting` as a predictor, shown in the prediction model below. The prediction model is basic prediction of TB incidence from subgroup and setting. As an example, I predicted the incidence for females and males in India (shown in the plot). The estimates for males (249.3660; CI = 139.68336, 359.0486) and females (123.1038; CI = 13.42112, 232.7864) are shown using `geom_point()` and `geom_errorbar()`. I need to explore the prediction modelling further before attempting to predict my main outcomes with more variables. 
```{r}
#prediction model for TB incidence with subgroup and setting as predictors
predinclm_fit <- lm_mod %>% 
  fit(`TB incidence (new infections per 100 000 population)` ~ subgroup + setting,
      data = highburden) #fit linear model
predincfittable <- tidy(predinclm_fit) 
print(predincfittable)

#prediction plot using the above model estimating TB incidence for females and males in India
readRDS(file = here("results", "model fit tables", "prediction1.rds")) 
```
## Full analysis
TBA: final analysis with statistic/ML methods, meaningful figures/tables, load saved products here
